Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-04-04 paper2.qmd
section:all; process:write,figures

<!-- 5. So, how do SCMs and DAGs support identification? -->
Luckily, SCMs and DAGs support identification analysis through two key advantages. First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks. This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@Neal_2020; @McElreath_2020]. Second, they depict causal relationships in a non-parametric way. This flexibility enables feasible identification strategies without requiring specification of the types of variables, the functional forms relating them, or the parameters of those functional forms [@Pearl_et_al_2016, pp. 35].

@@

<!-- 2. the preliminary SCM for CJ -->
[SCM -@fig-cj03_scm] illustrates the inclusion of the first theoretical principles into the conceptual-population model designed to examine the impact of different teaching methods on students' writing ability. This SCM outlines the relationship between the conceptual-population outcome $(O^{cp}_{hiabjk})$ and several related variables. The subscripts $h$ and $i$ label the students who authored the texts (individuals), while the indices $a$ and $b$ represent the compared texts (stimuli). The index $j$ refers to the judge, and $k$ represents the judgment index, accounting for experimental conditions where judges perform the same judgment multiple times, i.e., a *repeated measures design* [@Lawson_2015, pp. 366-376]. Notably, the indexing system facilitates comparisons between texts written by the same student $(h = i, a \neq b)$ and between texts written by distinct students ($h \neq i$, where $a = b$ is possible). However, it excludes cases where judges compare a student's text to itself $(h = i, a = b)$, as such comparison lacks practical relevance within the CJ framework.

@@

<!-- 1. Why is useful to declare the data structure? -->
Although specifying a data structure is not mandatory when using SCMs and DAGs, defining one in this case can improve clarity and facilitate the description of the system. Thus, to re-express the scalar form of the CJ system shown in @fig-cj03 into an equivalent vectorized form, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:

$$
I = \begin{bmatrix}
1 \\
2 \\
\vdots \\
h \\
i \\
\vdots \\
n_{I}
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
\vdots \\
j \\
\vdots \\
n_{J}
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{A} \\
\vdots & \vdots \\
h & a \\
\vdots & \vdots \\
i & b \\
\vdots & \vdots \\
n_{I} & 1 \\
\vdots & \vdots \\
n_{I} & n_{A} \\
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{K} \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & 1 \\
\vdots & \vdots \\
n_{J} & n_{K} \\
\end{bmatrix}
$$ {#eq-mat11}

<!-- 2. what is the data structure? -->
Here, each element of $I$ represents a unique individual $h$ or $i$, where $n_{I}$ denotes the total number of individuals. Similarly, each element of $J$ corresponds to a unique judge $j$, with $n_{J}$ indicating the total number of judges. Moreover, each row of $IA$ represents a unique pairing of individuals $h, i$ with stimuli $a, b$. As a result, the matrix $IA$ contains $n_{I} \cdot n_{A}$ rows and $2$ columns, where $n_{A}$ specifies the number of stimuli available per individual. Likewise, each row of $JK$ associates a judge $j$ with a judgment index $k$. Consequently, the matrix $JK$ has $n_{J} \cdot n_{K}$ rows and $2$ columns, where $n_{K}$ indicates the number of judgments each judge makes.

@@

<!-- $$ -->
<!-- {n_{I} \cdot n_{A} \choose 2} = \frac{ (n_{I} \cdot n_{A})! }{ 2! (n_{I} \cdot n_{A} - 2)! } -->
<!-- $$ -->

@@

It is easier to visualize the structure of these vectors and matrices by considering an example where $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$. In this simple case, the vectors and matrices described in equations ([-@eq-mat11]) and ([-@eq-mat12]) take the following form:

$$
I = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 \\
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
3 \\
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 \\
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 \\
\end{bmatrix} ; \;
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 1 & 1 \\
1 & 1 & 5 & 1 & 1 & 2 \\
1 & 1 & 5 & 1 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 3 & 1 \\
1 & 1 & 5 & 1 & 3 & 2 \\
1 & 1 & 5 & 1 & 3 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 \\
\end{bmatrix}
$$ {#eq-mat13}

@@

Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^3] judge-related variables $(Z_{J})$, such as judgment expertise, causally influence the latent variable representing judges' bias $(B_{J})$. It also shows how $B_{J}$ and *relevant* [^4] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes) causally influence the biases associated with each text $B_{JK}$, the second primary latent variable of interest. Additionally, the branch accounts for idiosyncratic errors: $e_{J}$ captures variations in judges' bias unexplained by $Z_{J}$, while $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$. Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$. Similarly, $Z_{JK}$ is an observed matrix with dimension $n_{J} \cdot n_{K}$ rows and $q_{JK}$ columns (variables), while $e_{JK}$ and $B_{JK}$ are latent vectors of length $n_{J} \cdot n_{K}$.

@@

<!-- 3. segway for next section -->
Overall, the conceptual-population model extends Thurstone's general form by introducing key innovations to address the limitations discussed in @sec-theory-issue1b and @sec-theory-issue2. These enhancements include accounting for judges' biases and integrating hierarchical structural components. Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to a *conceptual population* of comparative data, encompassing all repeated judgments from every judge for each pair of stimuli produced by every individual in the population. Since researchers rarely meet this assumption in practice, they must consider a more realistic scenario. The extensions to the conceptual-population model, derived from this realistic scenario, are described in the sample-comparison model.

@@

<!-- 1. What is the more realistic scenario: the sample mechanism -->
The sample-comparison model presents a more realistic scenario than the conceptual-population model. Expressly, we assume researchers work with a data sample consisting of a *limited number of judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$*, all drawn from the conceptual population. In a subsequent step, we assume that judges do not perform *all possible repeated comparisons* within the data sample. Instead, they conduct a sufficient number of comparisons, $n_{C}$, to ensure an accurate estimation of the proportion $P(B>A)$.

@@

<!-- 1. start by defining the sampling vector and matrices  -->
To incorporate the sampling mechanism and facilitate the interpretation of the sample-comparison model, we first define the *data sampling process* using the binary vector variables $S_{I}$, $S_{J}$, $S_{IA}$, and $S_{JK}$ as follows:

$$
S_{I} = \begin{bmatrix}
i_{(1)} \\
\vdots \\
i_{(h)} \\
\vdots \\
i_{(i)} \\
\vdots \\
i_{(nI)}
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
j_{(1)} \\
\vdots \\
j_{(j)} \\
\vdots \\
j_{(nJ)}
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
ia_{(1)} \\
\vdots \\
ia_{(ia)} \\
\vdots \\
ia_{(nI \cdot nA)} \\
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
jk_{(1)} \\
\vdots \\
jk_{(jk)} \\
\vdots \\
jk_{(nJ \cdot nK)} \\
\end{bmatrix}
$$ {#eq-mat21}

Where each element of $S_{I}$ and $S_{J}$ is a binary value indicating the presence or absence of data elements in the $I$ and $J$ vectors (see equation ([-@eq-mat11])). Thus, the vectors $S_{I}$ and $S_{J}$ contains $n_{I}$ and $n_{J}$ elements, respectively. Specifically, we define the vector $S_{I}$ as in equation ([-@eq-mat22]) and apply a similar definition to $S_{J}$.

$$
i_{(i)} = \begin{cases} 
1 & \text{if data element } i \text{ from } I \text{ is observed} \\
0 & \text{if data element } i \text{ from } I \text{ is missing}
\end{cases}
$$ {#eq-mat22}

Similarly, each element of $S_{IA}$ and $S_{JK}$ is a binary value indicating the presence or absence of data rows in the matrices $IA$ and $JK$ (see equation ([-@eq-mat11])). Thus, the vectors $S_{IA}$ and $S_{JK}$ contains $n_{I} \cdot n_{A}$ and $n_{J} \cdot n_{K}$ elements, respectively. Specifically, we define the vector $S_{IA}$ as in equation ([-@eq-mat23]) and apply a similar definition to $S_{JK}$.

$$
ia_{(ia)} = \begin{cases} 
1 & \text{if data elements } i,a \text{ from } IA \text{ are observed} \\
0 & \text{if data elements } i,a \text{ from } IA \text{ are missing}
\end{cases}
$$ {#eq-mat23}

We can visualize these vectors more clearly using the example in equation ([-@eq-mat13]). Suppose researchers exclude the second student and the third judge. Additionally, they do not include the second text from each individual and the third judgment from each judge. Given $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, these vectors take the following structure:

$$
S_{I} = \begin{bmatrix}
1 \\
0 \\
1 \\
1 \\
1
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
1 \\
1 \\
0
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
1 \\
0 \\
0 \\
0 \\
1 \\
0 \\
1 \\
0 \\
1 \\
0 
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
1 \\
1 \\
0 \\
1 \\
1 \\
0 \\
0 \\
0 \\
0 
\end{bmatrix}
$$ {#eq-mat24}

Equation ([-@eq-mat24]) shows that missing values in the $S_{I}$ and $S_{J}$ vectors directly determine those in the $S_{IA}$ and $S_{JK}$. In other words, researchers can only observe texts and judgments from students and judges initially included in the sample. Additionally, it shows that the sum of observed elements in $S_{I}$ equals the sampled students $(n^{s}_{I})$ and that a similar sum in vector $S_{J}$ equals the sampled judges $(n^{s}_{J})$. Conversely, the sum of observed elements in $S_{IA}$ represents the total sampled texts across all sampled students $(n^{s}_{I} \cdot n^{s}_{A})$, while a similar sum in vector $S_{JK}$ represents the total sampled judgments across all sampled judges $(n^{s}_{J} \cdot n^{s}_{K})$.

Finally, we define the *sample design* vector $S$ in equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every element of $S_{JK}$. Thus, the vector contains $n$ elements, matching the number of rows in the $R$ matrix, and the sum of its elements represents the total data sample: $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$. Here, the term ${n^{s}_{I} \cdot n^{s}_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of sampled stimuli generated by each pair of sampled individuals.

$$
S = \begin{bmatrix}
s_{(1)} \\
\vdots \\
s_{(s)} \\
\vdots \\
s_{(n)} \\
\end{bmatrix}
$$ {#eq-mat25}

<!-- 2. What is the effect of the sampling mechanism -->
With the definition of $S$, we incorporate the sample mechanism into the conceptual-population model. Following the convention of @McElreath_2020 [pp. 499-516] and @Deffner_et_al_2022 [pp. 6], the [DAG -@fig-cj12_dag] represents the conceptual-population outcome $O^{cp}_{R}$ as unobserved, emphasizing that researchers cannot directly access this outcome due to the sampling mechanism.

The DAG also depicts the *sample design* vector $S$ as a causal factor influencing the sample-comparison outcome $O^{sc}_{R}$. A square encloses this vector, indicating that it is a conditioned variable. In this context, *conditioning* means restricting the statistical analysis to a specific subset of the population based on the values of a conditioned variable [@Neal_2020, pp. 32], in this case $S$. Essentially, $S$ serves as the sample equivalent for the $R$ matrix (see equation ([-@eq-mat12])), that is, this vector selects *all repeated comparisons a subset of judges makes for the subset of stimuli produced by the sampled individuals*. This process ultimately generates the sample-comparison outcome $O^{sc}_{R}$ from the conceptual-population outcome $O^{cp}_{R}$.

@@

<!-- 1. start by defining the comparison vector  -->
As in the previous section, we first define the *comparison design* using the binary vector variable $C$ to facilitate the interpretation of the sample-comparison model and integrate the comparison mechanism.

$$
C = \begin{bmatrix}
c_{(1)} \\
\vdots \\
c_{(c)} \\
\vdots \\
c_{(n)} \\
\end{bmatrix}
$$ {#eq-mat26}

Where each element of $C$ is a binary value indicating the presence or absence of data rows in the $R$ matrix (see equation ([-@eq-mat12])):

$$
c_{c} = \begin{cases} 
1 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are observed} \\
0 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are missing}
\end{cases}
$$

Thus, the *comparison design* vector $C$ contains $n$ elements corresponding to the number of rows in the $R$ matrix, and the sum of its elements represents the selected comparisons, $n_{C}$. This definition assumes that judges do not perform *all possible repeated comparisons* within the data sample. Instead, they conduct a sufficient number of comparisons, $n_{C}$, to ensure the accurate estimation of the proportion $P(B>A)$ for each pair of stimuli [@Thurstone_1927b, pp. 267].

@@

In summary, extending Thurstone's general form with causal inference allows researchers to address BTL modeling issues, such as judge biases (see @sec-theory-issue1b), the hierarchical structure of stimuli (see @sec-theory-issue1b), and measurement errors in trait estimation and hypothesis testing (see @sec-theory-issue2). However, the model does not address concerns about the assumption of equal dispersion discussed in @sec-theory-issue1a. Since this issue involves statistical assumptions regarding the discriminal process distribution, it requires a statistical model, which we will develop in the next section.

>>>>

<!-- 5. So, how do SCMs and DAGs support identification? -->
Fortunately, SCMs and DAGs support identification analysis through two key advantages. First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [see @Neal_2020; @McElreath_2020]. This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis. Second, they depict causal relationships in a non-parametric way. This flexibility enables feasible identification strategies without requiring specification of the types of variables, the functional forms relating them, or the parameters of those functional forms [@Pearl_et_al_2016, pp. 35].

@@

<!-- 2. the preliminary SCM for CJ -->
[SCM -@fig-cj03_scm] presents the first theoretical principles embedded in the conceptual-population model, which evaluates the impact of different teaching methods on students' writing ability. This SCM outlines the relationship between the conceptual-population outcome $(O^{cp}_{hiabjk})$ and several related variables. The subscripts $h$ and $i$ identify the students who authored the texts (i.e., the individuals). The indices $a$ and $b$ represent the texts under comparison (i.e., the stimuli). The index $j$ indicates the judge conducting the comparison, while the index $k$ captures *repeated measures designs* [@Lawson_2015, pp. 366-376], accounting for experimental conditions where a judge compares the same pair of stimuli multiple times. Notably, the indexing system supports comparisons between different texts written by the same student $(h = i;$ $a \neq b)$ and between texts written by distinct students $(h \neq i;$ $a = b \; \text{is permitted})$, each judged once or repeatedly by all judges $(j = 1,\dots,n_{J};$ $k = 1,\dots,n_K)$. However, it excludes cases where a judge compares a student's text to itself, whether once or multiple times $(h = i;$ $a = b;$ $j = 1,\dots,n_{J};$ $k = 1,\dots,n_{K})$, as such comparison lacks practical relevance within the CJ framework. Here, $n_{J}$ indicates the total number of judges, and $n_{K}$ denotes the number of repeated judgments each judge performs.

@@

<!-- 1. Why is useful to declare the data structure? -->
Although specifying a data structure is not mandatory when using SCMs and DAGs, defining one in this case can improve clarity and facilitate the description of the system. Thus, to re-express the scalar form of the CJ system shown in @fig-cj03 into an equivalent vectorized form, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:
$$
I = \begin{bmatrix}
1 \\
2 \\
\vdots \\
h \\
\vdots \\
i \\
\vdots \\
n_{I}
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
\vdots \\
j \\
\vdots \\
n_{J}
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{A} \\
\vdots & \vdots \\
h & a \\
\vdots & \vdots \\
i & b \\
\vdots & \vdots \\
n_{I} & 1 \\
\vdots & \vdots \\
n_{I} & n_{A} \\
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{K} \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & 1 \\
n_{J} & 2 \\
\vdots & \vdots \\
n_{J} & n_{K} \\
\end{bmatrix}
$$ {#eq-mat11}

<!-- 2. what is the data structure? -->
Here, each element of $I$ represents a unique individual $h$ or $i$, where $n_{I}$ denotes the total number of individuals. Similarly, each element of $J$ corresponds to a unique judge $j$, with $n_{J}$ indicating the total number of judges. Moreover, each row of $IA$ represents a unique pairing of individuals $h, i$ with stimuli $a, b$. As a result, the matrix $IA$ contains $n_{I} \cdot n_{A}$ rows and $2$ columns, where $n_{A}$ specifies the number of stimuli available per individual. Likewise, each row of $JK$ associates a judge $j$ with a judgment index $k$. Consequently, the matrix $JK$ has $n_{J} \cdot n_{K}$ rows and $2$ columns, where $n_{K}$ indicates the number of repeated judgments each judge makes.

@@

@@

It is easier to visualize the structure of these vectors and matrices by considering an example where $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$. In this simple case, the vectors and matrices described in equations ([-@eq-mat11]) and ([-@eq-mat12]) take the following form:
$$
I = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 \\
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
3 \\
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 \\
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 \\
\end{bmatrix} ; \;
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 1 & 1 \\
1 & 1 & 5 & 1 & 1 & 2 \\
1 & 1 & 5 & 1 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 3 & 1 \\
1 & 1 & 5 & 1 & 3 & 2 \\
1 & 1 & 5 & 1 & 3 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 \\
\end{bmatrix}
$$ {#eq-mat13}

@@

Similarly, the bottom branch of [DAG -@fig-cj09_dag] depicts the second component, where *relevant* [^3] judge-related variables $(Z_{J})$, such as judgment expertise, and judges' idiosyncratic errors $e_{J}$ causally influence the latent variable representing judges' bias $(B_{J})$. The error $e_{J}$ captures variations in judges' bias unexplained by $Z_{J}$. This branch also shows how $B_{J}$, along with *relevant* [^4] judgment-related variables $(Z_{JK})$ (e.g., the number of judgments a judge makes), and judgments' idiosyncratic errors $e_{JK}$ causally influence the biases associated with each text $B_{JK}$, the second primary latent variable of interest. The error $e_{JK}$ captures variations in judgments unexplained by $B_{J}$ or $Z_{JK}$. Here, $Z_{J}$ is an observed matrix with $n_{J}$ rows and $q_{J}$ columns (variables), and both $e_{J}$ and $B_{J}$ are latent vectors of length $n_{J}$. Similarly, $Z_{JK}$ is an observed matrix with dimension $n_{J} \cdot n_{K}$ rows and $q_{JK}$ columns (variables), while $e_{JK}$ and $B_{JK}$ are latent vectors of length $n_{J} \cdot n_{K}$.

@@

<!-- 3. segway for next section -->
Overall, the conceptual-population model extends Thurstone's general form by introducing key innovations to address the limitations discussed in @sec-theory-issue1b and @sec-theory-issue2. These enhancements include accounting for judges' biases and integrating hierarchical structural components. Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to a *conceptual population* of comparative data. Since researchers rarely meet this assumption in practice, they must consider a more realistic scenario. The extensions to the conceptual-population model, derived from this realistic scenario, are described in the sample-comparison model.

@@

<!-- 1. What is the more realistic scenario: the sample mechanism -->
The sample-comparison model presents a more realistic scenario than the conceptual-population model. Expressly, it assumes researchers work with a data sample consisting of a limited number of repeated judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$, all drawn from the conceptual population. In a subsequent step, it assumes that judges do not perform *all possible stimuli comparisons* within the data sample. Instead, they conduct a sufficient number of stimuli comparisons, $n_{C}$, to ensure an accurate estimation of the proportion $P(B>A)$.

@@

<!-- 1. start by defining the sampling vector and matrices  -->
To incorporate the sampling mechanism and facilitate the interpretation of the sample-comparison model, we first define the *data sampling process* using the binary vector variables $S_{I}$, $S_{J}$, $S_{IA}$, and $S_{JK}$ as follows:
$$
S_{I} = \begin{bmatrix}
i_{(1)} \\
\vdots \\
i_{(h)} \\
\vdots \\
i_{(i)} \\
\vdots \\
i_{(nI)}
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
j_{(1)} \\
\vdots \\
j_{(j)} \\
\vdots \\
j_{(nJ)}
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
ia_{(1)} \\
\vdots \\
ia_{(ia)} \\
\vdots \\
ia_{(nI \cdot nA)} \\
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
jk_{(1)} \\
\vdots \\
jk_{(jk)} \\
\vdots \\
jk_{(nJ \cdot nK)} \\
\end{bmatrix}
$$ {#eq-mat21}

Where each element of $S_{I}$ and $S_{J}$ is a binary value indicating the presence or absence of data elements in the $I$ and $J$ vectors (see equation ([-@eq-mat11])). Thus, the vectors $S_{I}$ and $S_{J}$ contains $n_{I}$ and $n_{J}$ elements, respectively. Specifically, we define the vector $S_{I}$ as in equation ([-@eq-mat22]) and apply a similar definition to $S_{J}$.
$$
i_{(i)} = \begin{cases} 
1 & \text{if data element } i \text{ from } I \text{ is observed} \\
0 & \text{if data element } i \text{ from } I \text{ is missing}
\end{cases}
$$ {#eq-mat22}

Similarly, each element of $S_{IA}$ and $S_{JK}$ is a binary value indicating the presence or absence of data rows in the matrices $IA$ and $JK$ (see equation ([-@eq-mat11])). Thus, the vectors $S_{IA}$ and $S_{JK}$ contains $n_{I} \cdot n_{A}$ and $n_{J} \cdot n_{K}$ elements, respectively. Specifically, we define the vector $S_{IA}$ as in equation ([-@eq-mat23]) and apply a similar definition to $S_{JK}$.
$$
ia_{(ia)} = \begin{cases} 
1 & \text{if data elements } i,a \text{ from } IA \text{ are observed} \\
0 & \text{if data elements } i,a \text{ from } IA \text{ are missing}
\end{cases}
$$ {#eq-mat23}

We can visualize these vectors more clearly using the example in equation ([-@eq-mat13]). Suppose researchers exclude the second student and the third judge. Additionally, they do not include the second text from each individual and the third judgment from each judge. Given $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, these vectors take the following structure: 
$$
S_{I} = \begin{bmatrix}
1 \\
0 \\
1 \\
1 \\
1
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
1 \\
1 \\
0
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
1 \\
0 \\
0 \\
0 \\
1 \\
0 \\
1 \\
0 \\
1 \\
0 
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
1 \\
1 \\
0 \\
1 \\
1 \\
0 \\
0 \\
0 \\
0 
\end{bmatrix}
$$ {#eq-mat24}

Equation ([-@eq-mat24]) shows that missing values in the $S_{I}$ and $S_{J}$ vectors directly determine those in the $S_{IA}$ and $S_{JK}$. In other words, researchers can only observe texts and judgments from students and judges initially included in the sample. Additionally, it shows that the sum of observed elements in $S_{I}$ equals the sampled students $(n^{s}_{I})$ and that a similar sum in vector $S_{J}$ equals the sampled judges $(n^{s}_{J})$. Conversely, the sum of observed elements in $S_{IA}$ represents the total sampled texts across all sampled students $(n^{s}_{I} \cdot n^{s}_{A})$, while a similar sum in vector $S_{JK}$ represents the total sampled repeated judgments across all sampled judges $(n^{s}_{J} \cdot n^{s}_{K})$.

Finally, we define the *sample mechanism* vector $S$ in equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every element of $S_{JK}$. Thus, the vector contains $n$ elements, matching the number of rows in the $R$ matrix, and the sum of its elements represents the total data sample: $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$. Here, the term ${n^{s}_{I} \cdot n^{s}_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of sampled stimuli generated by each pair of sampled individuals.
$$
S = \begin{bmatrix}
s_{(1)} \\
\vdots \\
s_{(s)} \\
\vdots \\
s_{(n)} \\
\end{bmatrix}
$$ {#eq-mat25}

<!-- 2. What is the effect of the sampling mechanism -->
With the definition of $S$, we incorporate the sample mechanism into the conceptual-population model. Following the convention of @McElreath_2020 [pp. 499-516] and @Deffner_et_al_2022 [pp. 6], [DAG -@fig-cj12_dag] represents the conceptual-population outcome $O^{cp}_{R}$ as unobserved, emphasizing that researchers cannot directly access this outcome due to the sampling mechanism. The DAG also depicts the *sample design* vector $S$ as a causal factor influencing the sample-comparison outcome $O^{sc}_{R}$. A square encloses this vector, indicating that it is a conditioned variable. In this context, *conditioning* means restricting the statistical analysis to a specific subset of the population based on the values of a conditioned variable [@Neal_2020, pp. 32], in this case $S$. Essentially, $S$ serves as the sample equivalent for the $R$ matrix (see equation ([-@eq-mat12])), that is, this vector selects *all repeated judgments a subset of judges makes for the subset of stimuli produced by the sampled individuals*. This process ultimately generates the sample-comparison outcome $O^{sc}_{R}$ from the conceptual-population outcome $O^{cp}_{R}$.

@@

<!-- 1. start by defining the comparison vector  -->
As in the previous section, to facilitate the interpretation of the sample-comparison model, we first define the *comparison mechanism* using the binary vector variable $C$ as follows: 
$$
C = \begin{bmatrix}
c_{(1)} \\
\vdots \\
c_{(c)} \\
\vdots \\
c_{(n)} \\
\end{bmatrix}
$$ {#eq-mat26}

Where $C$ contains $n$ elements corresponding to the number of rows in the $R$ matrix, with each element being a binary value indicating the presence or absence of data rows in the $R$ matrix (see equation ([-@eq-mat12])): 
$$
c_{(c)} = \begin{cases} 
1 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are observed} \\
0 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are missing}
\end{cases}
$$

This definition assumes that the subset of judges *do not* perform all repeated judgments for the subset of stimuli produced by the sampled individuals. Instead, they conduct a "sufficient number of comparisons", $n_{C}$, to ensure the accurate estimation of the proportion $P(B>A)$ for each pair of stimuli [@Thurstone_1927b, pp. 267].

@@

In summary, extending Thurstone's general form using Causal inference allows researchers to address BTL modeling issues, such as judge biases (see @sec-theory-issue1b), the hierarchical structure of stimuli (see @sec-theory-issue1b), and measurement errors in trait estimation and hypothesis testing (see @sec-theory-issue2). However, the model does not address concerns about the assumption of equal dispersion discussed in @sec-theory-issue1a. Since this issue involves statistical assumptions regarding the discriminal process distribution, it requires a statistical model, which we will develop in the next section.
